KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-05-16 paper_overleaf/KIN_a_method_to_estimate_relatedness_from_low_coverage_data_supplements.tex

\begin{figure}[h!]
    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
    \centering
    \caption{Comparison of beta distributions estimated with the Identical KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows for identical individuals. Colored lines shows the Beta probability distribution for each IBD state $Z_w$.}
    \label{figS4:bndsbeta}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
    \caption{Comparison of IBD estimation in control simulations and in presence of contamination, ascertainment or ROH (denoted in figure legend by Control, C, A and R, respectively) at different coverages. y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario (control, C, A or R), and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
    \label{figS11:ibd_others}
\end{figure}

>>>>

\begin{figure}[h!]
    \includegraphics[width=18cm]{supplementary_info/plots/contam0_inbred1_run57_coverage0.2_asc0_inputMode_hapProbs_fil0_pair0_15_relid_betaplot.png}
    \centering
    \caption{Comparison of beta distributions estimated with the Identical KIN-HMM (left) without and (right) with variance constrained optimization of $\delta$ parameters.The histogram in each plot shows pairwise differences in genomic windows for identical individuals. Colored lines shows the Beta probability distribution for each $i$, where $i$ is the index used for combinations of $Z_W$ and $H_w$.}
    \label{figS4:bndsbeta}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=18cm]{plots/plotimg/plot_IBDaccuracy_others_supp.png}
    \caption{Comparison of IBD estimation in control simulations and in presence of contamination, ascertainment or ROH (denoted in figure legend by Control, C, A and R, respectively) at different coverages. The y-axis shows accuracy calculated over 60 simulations for each relatedness case, for each scenario (control, C, A or R), and six different coverages shown on the x-axis. The error bars display one standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other.}
    \label{figS11:ibd_others}
\end{figure}

### 2022-05-16 paper_overleaf/main.tex

\paragraph{ROH detection:}
We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}). We computed the proportion of ROH in all windows and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows. Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}). With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.004$). The largest bias observed ($9\%$) was for the datasets that included contamination. Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}). Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz. proportion of window in ROH, and this state is inferred more confidently with more data. Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).

@@

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations does not affect IBD prediction. Adding ROH to the simulations does not affect identical, parent-child, unrelated and $3^{rd} degree$  pairs, but reduces the IBD prediction accuracy for second degree by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see figure \ref{fig3:Comparison_READ_KIN})

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin. We found that in most cases the three methods agree. Only $<5\%$ of pairs (315 out of 6903) show a different relatedness for KIN and READ. However, most of these are at very low confidence for both the methods ($\Delta LL<1$ for KIN and $sd<1$ for READ). There are 28 pairs where KIN shows confident results (table S2), and when we plot the raw heterozygosity estimates along the genome, in many cases it is possible to see that the prediction by KIN is in agreement with the observed proportion of differences along the genome (Fig. S11). Comparisons with lcMLkin are sparse since lcMLkin shows reliable results for $\appr 53\%$ pairs (\cite{mittnik_kinship-based_2019}), and we found 16 pairs for which lcMLkin gives a reliable result, and KIN shows $\Delta LL>$1 (table S3). We again investigated these comparisons in more detail, and found that at least in cases where there is good amount of overlapping sites ($\geq35,000$), we can clearly see that the proportion of differences along the genome aligns with the prediction of KIN (Fig. S12).

@@

\paragraph{Simulations for method evaluation:}
Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (Fig. S1). We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively.  We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1. This scenario roughly models the ascertainment of the Chagyrskaya and Okladnikov Caves data. We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity in unrelated individuals with a Markov chain using the transition matrix:

@@

\section{Acknowledgements}
We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo Iasi, Alba Bossom Mesa, Arev Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions. This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.

\section{Contributions}
Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: B.M.P.; Methodology—support: D.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing (original draft preparation)—lead: D.P.; Writing (original draft preparation)—support: B.M.P.; Writing (review and editing): input from all authors; Supervision: B.M.P.

>>>>

\paragraph{ROH detection}
We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}). We computed the proportion of ROH in all windows and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows. Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}). With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.004$). The largest bias observed ($9\%$) was for the datasets that included contamination. Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}). Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz. proportion of window in ROH, and this state is inferred more confidently with more data. Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).

@@

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatdness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see figure \ref{fig3:Comparison_READ_KIN})

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech Valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained by the authors using READ and lcMLkin. We found that KIN was able to confidently classify $85\%$ of the 6903 possible comparisons ($\Delta LL<1$), READ $94\%$ ($sd<1$) and lcMLkin 53\% of pairs (\cite{mittnik_kinship-based_2019}).

Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, figure S11). For 20 of these comparisons, KIN predicts a third-degree relationship, which READ classifies as unrelated.

lcMLkin predicts $3^{rd}$ - $5^th$ in 14 of these cases, unrelated in 1 case, second degree in 1 case, and does not have enough data for classification in 4 cases. Since READ does not infer third degree, and lcMLkin infers a group of relatedness together as $3^{rd}$ - $5^th$, it is likely that KIN is able to correctly resolve these cases. For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^th$ degree, but KIN infers them to be unrelated. Thus, these are likely examples of $4^{th}$ or $5^{th}$ degree relations.

For less than third-degree relations, only eight cases are classified differently between KIN and READ. lcMLkin matches KIN's prediction in two cases, and does not have enough data in 5 cases. In the last case, all three methods differ (table S3). Finally, there are three disagreements between KIN and lcMLkin in classification of parent-child versus siblings (READ predicts first degree for all three pairs). We plotted the pairwise differences for these three pairs in Fig. S12, and found that the proportion of differences along the genome aligns with the prediction of KIN.

@@

\paragraph{Simulations for method evaluation:}
Apart from the related and unrelated individuals in Pop1, we simulated more haploid individuals in three other populations to create scenarios with ascertainment bias and contamination (Fig. S1). We simulated two individuals to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 2000 generations and 4000 generations ago respectively.  We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1. This scenario roughly models the ascertainment of the Chagyrskaya and Okladnikov Caves data. We tested the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity in unrelated individuals with a Markov chain using the transition matrix:

@@

\section{Acknowledgements}
We thank Svante Pääbo, Janet Kelso, Harald Ringbauer, Johann Visagie, Laurits Skov, Leonardo N. M. Iasi, Alba Bossoms Mesa, Arev P. Sümer, Zuzana Hofmanová, Guido Alberto Gnecchi Ruscone, Ke Wang and Luca Traverso for helpful comments and discussions. This work was supported by the Max Planck Society and the European Research Council (Grant No.: 694707) to Svante Pääbo.

\section{Contributions}
Conceptualization (Design of study): B.M.P.; Software: D.P.; Methodology—lead: D.P.; Methodology—support: B.M.P., S.P.; Formal Analysis: D.P.; Visualization-lead: D.P.; Visualization-support: S.P.; Data Curation: D.P.; Writing (original draft preparation)—lead: D.P.; Writing (original draft preparation)—support: B.M.P.; Writing (review and editing):  all authors; Supervision: B.M.P.
